---
title: South Okanagan Bat Hibernacula Project 2019-2025
subtitle: 2025-26 Update
format:
  html:
    grid:
      margin-width: 300px
    title-block-banner: "#1d4566"
    title-block-banner-color: "#F4F4F5"
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-expand: true
    toc-location: left
    css: styles.css
  pdf:
    tbl-pos: "H"
    toc: true
    toc-depth: 3
    toccolor: black
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
    papersize: letter
    extra_dependencies: ["float"]
    include-before-body:
      text: |
        \clearpage
    include-in-header:
      text: |
        \usepackage{placeins}
        \usepackage{float}
        \floatplacement{table}{H}
date: last-modified
author: 
  - name: "Mandy Kellner"
    affiliation: "Ministry of Water, Land and Resource Stewardship, British Columbia"
  - name: "Camila Hurtado"
    affiliation: "Biodiversity Pathways"
editor: visual
bibliography: NABat-NNWHub-references.bib
nocite: '@*'
---

```{r}
#| label: Load packages and data
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
 
library(tidyr)
library(dplyr)
library(lubridate)
library(stringr)
library(kableExtra)
library(tidyverse)
library(leaflet)
library(base64enc)
library(mapview)
library(sf)
library(gt)

OK <- read.csv("Data/merged_locations_aggregated2012-2025.csv")
sites <- read.csv("Data/Sites.csv")|>   # update path as needed
  # Rename columns to something workable
  rename(
    park      = Park...Protected.Area,
    site      = Site,
    lat       = Latitude,
    lon       = Longitude,
    y1112     = Years.monitored,  # 2011-12
    y1819     = X,                # 2018-19
    y1920     = X.1,                # 2019-20
    y2021     = X.2,              # 2020-21
    y2122     = X.3,              # 2021-22
    y2223     = X.4,              # 2022-23
    y2425     = X.5,              # 2024-25
    data_type = Data.type,
    source    = Project.report.data.source..BC.WSI.project.ID.6430.
  ) |>
  filter(site !="") |> # drop the header-artifact row (row 1 in your CSV)
  mutate(
  lat = formatC(lat, format = "f", digits = 4),
  lon = formatC(lon, format = "f", digits = 4))
#get values for results summary
# Extract year from Detection.Date
OK <- OK %>%
  mutate(Year = if_else(
    SampleYear == "2011-12",
    as.numeric(Detection.Date),
    as.numeric(year(as.Date(Detection.Date, format = "%Y-%m-%d")))
  ))

# --- Most recordings: site x year combo ---
site_year_counts <- OK %>%
  group_by(Sample.Station.Label, Year) %>%
  summarise(n_recordings = n(), .groups = "drop")

top_site_year <- site_year_counts %>%
  slice_max(n_recordings, n = 1, with_ties = FALSE)

top_site      <- top_site_year$Sample.Station.Label
top_site_yr   <- top_site_year$Year
top_site_n    <- top_site_year$n_recordings

# --- Most sites sampled: which year had the most unique sites ---
sites_per_syear <- OK %>%
  group_by(SampleYear) %>%
  summarise(n_sites = n_distinct(Sample.Station.Label), .groups = "drop")

top_sampled_yr     <- sites_per_syear %>% slice_max(n_sites, n = 1, with_ties = FALSE)
top_sampled_year   <- top_sampled_yr$SampleYear
top_sampled_nsites <- top_sampled_yr$n_sites

# ------------------------- Species summaries ------------------------
# Codes to exclude
exclude_codes <- c("NOISE", "NoID", "HIF", "LOF", "HILO")

spp_clean <- OK %>%
  filter(
    !Species.Code %in% exclude_codes,
    !is.na(Species.Code),
    Species.Code != ""
  )

# Most common species (by number of detections)
spp_counts <- spp_clean %>%
  group_by(Species.Code) %>%
  summarise(n = n(), .groups = "drop")

top_spp   <- spp_counts %>% slice_max(n, n = 1, with_ties = FALSE)
top_spp_code <- top_spp$Species.Code
top_spp_n    <- top_spp$n

# Total confirmed species: exactly 4 characters, after exclusions
confirmed_spp <- spp_clean %>%
  filter(nchar(as.character(Species.Code)) == 4) %>%
  distinct(Species.Code)

n_confirmed <- nrow(confirmed_spp)
```

```{r}
#| label: Prep data and create map
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

# ── 1. Summarise species per site ─────────────────────────────────────────────
exclude_ids <- c("Noise", "NoID", "NOISE", "NOID", "")

species_by_site_year <- OK %>%
  filter(!Species.Code %in% exclude_ids,
         !is.na(Species.Code)) %>%
  mutate(SYear = as.character(SampleYear),
         # For pie charts: couplets (e.g. EPFULANO) now stay as themselves so
         # they can be drawn as striped slices. Only group codes / HILO-HIF-LOF
         # still collapse into "Other".
         species_pie = case_when(
           toupper(Species.Code) %in% c("HILO", "HIF", "LOF")    ~ "Other",
           # Couplets/multiplets (>= 8 letters, e.g. EPFULANO) stay as themselves
           # so they can be drawn as striped slices -- even when a Bat Species
           # Group Code is populated on the row (which would otherwise collapse
           # them into "Other" via the rule below).
           nchar(Species.Code) >= 8 &
             grepl("^([A-Za-z]{4})+$", Species.Code)             ~ Species.Code,
           !is.na(Bat.Species.Group.Code) &
             Bat.Species.Group.Code != ""                         ~ "Other",
           .default = Species.Code
         ),
         # For tables: keep everything as-is
         species_table = Species.Code,
         # Project/source for the new project filter (NA/"" -> "Unknown")
         Source = { s <- as.character(Source); ifelse(is.na(s) | s == "", "Unknown", s) }
  ) %>%
  group_by(Sample.Station.Label, Longitude..DD., Latitude..DD., SYear, species_pie, species_table, Source) %>%
  summarise(n = sum(Count), .groups = "drop")

# ── 2. Build site coords ───────────────────────────────────────────────────────
map_data <- species_by_site_year %>%
  rename(site_name = Sample.Station.Label,
         lon = Longitude..DD.,
         lat = Latitude..DD.) %>%
  filter(!is.na(lat))

# ── 3. SVG pie helper ─────────────────────────────────────────────────────────
# Couplet slices (EPFULANO, MYCAMYYU, MYCIMYLUMYVO, ...) are filled with a
# diagonal stripe pattern built from the colours of their constituent species.
make_pie_svg <- function(values, colors, labels = NULL, color_lookup = NULL,
                         size = 70, uid = "pie") {
  total <- sum(values)
  if (total == 0) return(NULL)
  if (is.null(labels))       labels       <- rep("", length(values))
  if (is.null(color_lookup)) color_lookup <- character(0)
  fracs  <- values / total
  angles <- cumsum(c(0, fracs)) * 2 * pi
  r <- size / 2 - 2
  
  esc_attr <- function(s) {
    s <- gsub("&", "&amp;",  s, fixed = TRUE)
    s <- gsub('"', "&quot;", s, fixed = TRUE)
    s <- gsub("<", "&lt;",   s, fixed = TRUE)
    s <- gsub(">", "&gt;",   s, fixed = TRUE)
    s
  }
  
  # A code is a couplet/multiplet if it is >= 8 chars, a multiple of 4, and
  # every 4-char chunk is all letters (e.g. EPFULANO -> EPFU, LANO).
  couplet_parts <- function(code) {
    if (is.na(code) || nchar(code) < 8 || nchar(code) %% 4 != 0) return(NULL)
    starts <- seq(1, nchar(code), by = 4)
    parts  <- substring(code, starts, starts + 3)
    if (!all(grepl("^[A-Za-z]{4}$", parts))) return(NULL)
    parts
  }
  
  defs  <- character(0)
  fills <- character(length(values))
  for (i in seq_along(values)) {
    parts <- couplet_parts(labels[i])
    if (!is.null(parts)) {
      cols <- vapply(toupper(parts), function(p)
        if (p %in% names(color_lookup)) color_lookup[[p]] else "#AAAAAA",
        character(1))
      pid <- sprintf("stripe_%s_%d", uid, i)
      sw  <- 7                       # stripe width (px)
      pw  <- sw * length(cols)       # one full colour cycle
      rects <- paste(vapply(seq_along(cols), function(k) sprintf(
        '<rect x="%s" y="0" width="%s" height="%s" fill="%s"/>',
        (k - 1) * sw, sw, size * 2, cols[k]), character(1)), collapse = "")
      defs <- c(defs, sprintf(
        '<pattern id="%s" patternUnits="userSpaceOnUse" width="%s" height="%s" patternTransform="rotate(45)">%s</pattern>',
        pid, pw, size * 2, rects))
      fills[i] <- sprintf("url(#%s)", pid)
    } else {
      fills[i] <- colors[i]
    }
  }
  
  slices <- purrr::pmap_chr(
    list(i = seq_along(values), fl = fills, lab = labels, val = values, fr = fracs),
    function(i, fl, lab, val, fr) {
      dl <- if (nzchar(lab))
        sprintf(' data-label="%s: %s (%s%%)"', esc_attr(lab), val, round(fr * 100, 1)) else ''
      a0 <- angles[i]; a1 <- angles[i + 1]
      if (abs(a1 - a0) >= 2 * pi - 1e-6) {
        sprintf('<circle cx="%s" cy="%s" r="%s" fill="%s"%s/>',
                size/2, size/2, r, fl, dl)
      } else {
        x0 <- size/2 + r*sin(a0); y0 <- size/2 - r*cos(a0)
        x1 <- size/2 + r*sin(a1); y1 <- size/2 - r*cos(a1)
        large <- if ((a1 - a0) > pi) 1L else 0L
        sprintf('<path d="M %s %s L %s %s A %s %s 0 %s 1 %s %s Z" fill="%s" stroke="white" stroke-width="1"%s/>',
                size/2, size/2, x0, y0, r, r, large, x1, y1, fl, dl)
      }
    }
  )
  defs_block <- if (length(defs)) sprintf("<defs>%s</defs>", paste(defs, collapse = "")) else ""
  sprintf('<svg xmlns="http://www.w3.org/2000/svg" width="%s" height="%s">%s%s</svg>',
          size, size, defs_block, paste(slices, collapse = ""))
}

# ── 4. Colour palette ─────────────────────────────────────────────────────────
# Grouped by ECHOLOCATION CALL FREQUENCY (not detection count):
#   - High-frequency callers (Myotis spp.) -> warm reds -> yellows
#   - Low-frequency callers                 -> cool blues + greens
# Within each group the hue tracks detection abundance (darkest/most saturated =
# most detected), using the % noted in each comment.
sp_colors <- c(
  # -- high-frequency callers (Myotis spp.): reds -> yellows ------------------
  "MYCI"  = "#B10026",   # 16.8%  dark red
  "MYCA"  = "#E31A1C",   # 4.9%   red
  "MYYU"  = "#FC4E2A",   # 2.4%   red-orange
  "MYLU"  = "#FD8D3C",   # 2.0%   orange
  "MYTH"  = "#FEB24C",   # 0.4%   light orange
  "MYVO"  = "#FED976",   # 0.2%   yellow
  "MYEV"  = "#FFEDA0",   # 0.1%   pale yellow
  # -- low-frequency callers: blues + greens ----------------------------------
  "LANO"  = "#08519C",   # 13.9%  dark blue
  "EPFU"  = "#006D2C",   # 4.7%   dark green
  "ANPA"  = "#2171B5",   # 1.5%   blue
  "EUMA"  = "#238B45",   # 0.9%   green
  "COTO"  = "#4292C6",   # 0.5%   medium blue
  "PAHE"  = "#41AB5D",   # 0.1%   medium green
  "LACI"  = "#6BAED6",   # 6 obs  light blue
  "LABO"  = "#74C476",   # 3 obs  light green
  "Other" = "#AAAAAA"    #        grey
)

# ── 5. Helper: build one HTML table body ──────────────────────────────────────
make_sp_table <- function(sp_df, pie_colors) {
  sp_df %>%
    arrange(desc(n)) %>%
    mutate(pct = round(n / sum(n) * 100, 1)) %>%
    purrr::pmap_chr(function(species, n, pct) sprintf(
      '<tr><td><span style="display:inline-block;width:10px;height:10px;border-radius:50%%;background:%s;margin-right:4px"></span>%s</td><td align="right">%s</td><td align="right">%s%%</td></tr>',
      ifelse(species %in% names(pie_colors), pie_colors[species], "#AAAAAA"),
      species, n, pct
    )) %>%
    paste(collapse = "")
}

make_sp_header <- function(uid) {
  sprintf(
    '<tr>
       <th align="left"  style="cursor:pointer;user-select:none" onclick="window.sortSpeciesTable(\'%s\',0,\'text\')">Species &#8597;</th>
       <th align="right" style="cursor:pointer;user-select:none" onclick="window.sortSpeciesTable(\'%s\',1,\'num\')">n &#8597;</th>
       <th align="right" style="cursor:pointer;user-select:none" onclick="window.sortSpeciesTable(\'%s\',2,\'num\')">%% &#8597;</th>
     </tr>',
    uid, uid, uid
  )
}

# ── 5b. make_sp_table: species code and numbers only, no colour swatch ────────
# Row order: 4-letter codes first, then any other text codes, then codes that
# begin with a digit. Alphabetical within each group.
sp_code_tier <- function(x) {
  dplyr::case_when(
    grepl("^[0-9]", x) ~ 3L,   # numeric-leading codes go last
    nchar(x) == 4      ~ 1L,   # standard 4-letter codes first
    .default = 2L              # everything else in between
  )
}

make_sp_table <- function(sp_df, pie_colors = NULL) {
  sp_df %>%
    mutate(
      pct   = round(n / sum(n) * 100, 1),
      .tier = sp_code_tier(species),
      .key  = toupper(species)
    ) %>%
    arrange(.tier, .key) %>%
    select(species, n, pct) %>%
    purrr::pmap_chr(function(species, n, pct) sprintf(
      '<tr><td>%s</td><td align="right">%s</td><td align="right">%s%%</td></tr>',
      species, n, pct
    )) %>%
    paste(collapse = "")
}

# ── 6. Build site list ────────────────────────────────────────────────────────
site_list <- map_data %>%
  group_by(site_name, lat, lon) %>%
  summarise(
    sp_data = list(tibble(species_pie, species_table, n, SYear, Source)),
    total   = sum(n),
    years   = list(sort(unique(SYear))),
    .groups = "drop"
  )

# ── 6b. Colour lookup for the FILTERED view only ──────────────────────────────
all_species_codes <- sort(unique(species_by_site_year$species_table))
unknown_codes     <- setdiff(all_species_codes, names(sp_colors))
fallback_codes    <- if (length(unknown_codes)) {
  setNames(colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(unknown_codes)),
           unknown_codes)
} else character(0)
species_colors_all <- c(sp_colors, fallback_codes)

# -- Sources (projects) for the project filter: stable list + one colour each --
all_sources   <- sort(unique(map_data$Source))
source_pal    <- c("#4E79A7", "#F28E2B", "#59A14F", "#B07AA1", "#9C755F", "#EDC948")
source_colors <- setNames(source_pal[((seq_along(all_sources) - 1) %% length(source_pal)) + 1],
                          all_sources)

# ── 7. Assemble one payload per site (no markers added in R) ──────────────────
# Slice rule: any species that is >= 2% of the site's total is drawn as its own
# slice; everything below 2% (plus the pre-existing "Other") is pooled into Other.
pie_min_share <- 0.02

sites_payload  <- vector("list", nrow(site_list))
all_pie_colors <- c()

for (i in seq_len(nrow(site_list))) {
  row   <- site_list[i, ]
  sp    <- row$sp_data[[1]]
  years <- row$years[[1]]
  
  if (nrow(sp) == 0 || sum(sp$n) == 0) next
  
  # -- Pie chart categories, all years -----------------------------------------
  sp_pie_all <- sp %>%
    group_by(species_pie) %>%
    summarise(n = sum(n), .groups = "drop") %>%
    arrange(desc(n)) %>%
    rename(species = species_pie)
  
  site_total_pie <- sum(sp_pie_all$n)
  keep_species <- sp_pie_all %>%
    filter(species != "Other", n / site_total_pie >= pie_min_share) %>%
    pull(species)
  
  sp_pie <- sp_pie_all %>%
    mutate(species = if_else(species %in% keep_species, species, "Other")) %>%
    group_by(species) %>%
    summarise(n = sum(n), .groups = "drop") %>%
    arrange(desc(n))
  
  # -- Pie colours: use the SINGLE global map (species_colors_all) so an unknown
  #    code (e.g. Myotis, 40KMyo) gets the SAME colour at every site. Couplets
  #    are striped, so the solid colour here is only a placeholder that
  #    make_pie_svg ignores. -----------------------------------------------------
  pie_colors <- setNames(
    vapply(sp_pie$species, function(s)
      if (s %in% names(species_colors_all)) species_colors_all[[s]] else "#AAAAAA",
      character(1)),
    sp_pie$species)
  
  # -- Table colours: raw species codes, per-site fallback ---------------------
  all_sp       <- unique(sp$species_table)
  unknown_all  <- setdiff(all_sp, names(sp_colors))
  fallback_all <- if (length(unknown_all)) {
    setNames(colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(length(unknown_all)),
             unknown_all)
  } else character(0)
  all_colors <- c(sp_colors, fallback_all)
  
  all_pie_colors <- c(all_pie_colors, pie_colors)
  
  uid     <- gsub("[^a-zA-Z0-9]", "_", row$site_name)
  pie_svg <- make_pie_svg(sp_pie$n, pie_colors[sp_pie$species],
                          labels = sp_pie$species,
                          color_lookup = species_colors_all, uid = uid)
  
  # -- All-years table: raw species codes --------------------------------------
  sp_all_table <- sp %>%
    group_by(species_table) %>%
    summarise(n = sum(n), .groups = "drop") %>%
    rename(species = species_table)
  
  all_table_html <- paste0(
    make_sp_header(uid),
    make_sp_table(sp_all_table, all_colors)
  )
  
  # -- Tab buttons + per-year tables -------------------------------------------
  year_button_list <- c(
    sprintf(
      '<button onclick="(function(b){var uid=\'%s\';var rows=b.getAttribute(\'data-rows\');var tot=b.getAttribute(\'data-tot\');document.getElementById(\'tbl_\'+uid).innerHTML=rows;document.getElementById(\'lbl_\'+uid).innerHTML=\'All years &mdash; \'+tot+\' detections\';[].forEach.call(document.querySelectorAll(\'[data-uid%s]\'),function(x){x.style.background=\'#e0e0e0\';x.style.color=\'#333\'});b.style.background=\'#4E79A7\';b.style.color=\'white\';})(this)" data-uid%s="1" data-tot="%s" data-rows="%s" id="btn_%s_all" style="margin:2px;padding:3px 8px;border:1px solid #ccc;border-radius:4px;background:#4E79A7;color:white;cursor:pointer;font-size:11px;">All</button>',
      uid, uid, uid, row$total,
      htmltools::htmlEscape(all_table_html, attribute = TRUE),
      uid
    ),
    purrr::map_chr(years, function(yr) {
      sp_yr <- sp %>%
        filter(SYear == yr) %>%
        group_by(species_table) %>%
        summarise(n = sum(n), .groups = "drop") %>%
        rename(species = species_table)
      yr_total <- sum(sp_yr$n)
      yr_table_html <- paste0(
        make_sp_header(uid),
        make_sp_table(sp_yr, all_colors)
      )
      sprintf(
        '<button onclick="(function(b){var uid=\'%s\';var rows=b.getAttribute(\'data-rows\');var tot=b.getAttribute(\'data-tot\');var yr=b.getAttribute(\'data-yr\');document.getElementById(\'tbl_\'+uid).innerHTML=rows;document.getElementById(\'lbl_\'+uid).innerHTML=yr+\' &mdash; \'+tot+\' detections\';[].forEach.call(document.querySelectorAll(\'[data-uid%s]\'),function(x){x.style.background=\'#e0e0e0\';x.style.color=\'#333\'});b.style.background=\'#4E79A7\';b.style.color=\'white\';})(this)" data-uid%s="1" data-yr="%s" data-tot="%s" data-rows="%s" id="btn_%s_%s" style="margin:2px;padding:3px 8px;border:1px solid #ccc;border-radius:4px;background:#e0e0e0;color:#333;cursor:pointer;font-size:11px;">%s</button>',
        uid, uid, uid, yr, yr_total,
        htmltools::htmlEscape(yr_table_html, attribute = TRUE),
        uid, yr, yr
      )
    })
  ) %>% paste(collapse = "")
  
  n_species <- nrow(sp_all_table)
  table_style <- if (n_species > 15) {
    'style="width:100%;margin-top:4px;display:block;max-height:250px;overflow-y:auto;"'
  } else {
    'style="width:100%;margin-top:4px;"'
  }
  
  popup <- sprintf(
    '<div style="font-family:sans-serif;min-width:220px;max-width:280px">
  <b>%s</b>
  <hr style="margin:5px 0"/>
  <div style="margin-bottom:6px">%s</div>
  <b id="lbl_%s">All years &mdash; %s detections</b>
  <table id="tbl_%s" %s>%s</table>
</div>',
    row$site_name, year_button_list, uid, row$total, uid, table_style, all_table_html
  )
  
  # -- species x year counts, for the filtered popups --------------------------
  sp_years <- sp %>%
    group_by(species_table, SYear) %>%
    summarise(n = sum(n), .groups = "drop") %>%
    arrange(species_table, SYear)
  
  species_list <- unname(purrr::imap(
    split(sp_years, sp_years$species_table),
    function(d, code) list(
      code   = code,
      total  = sum(d$n),
      years  = I(as.character(d$SYear)),   # I() stops jsonlite unboxing length-1
      counts = I(as.integer(d$n))
    )
  ))
  
  # -- per-source (project) species breakdown, for the project filter ----------
  src_species <- sp %>%
    group_by(Source, species_table) %>%
    summarise(n = sum(n), .groups = "drop") %>%
    arrange(Source, desc(n))
  
  sources_list <- unname(purrr::imap(
    split(src_species, src_species$Source),
    function(d, nm) list(
      name   = nm,
      total  = sum(d$n),
      codes  = I(as.character(d$species_table)),
      counts = I(as.integer(d$n))
    )
  ))
  
  sites_payload[[i]] <- list(
    site    = row$site_name,
    lat     = row$lat,
    lon     = row$lon,
    total   = row$total,
    icon    = pie_svg,
    popup   = popup,
    species = species_list,
    sources = sources_list
  )
}

sites_payload <- Filter(Negate(is.null), sites_payload)

# ── 8. JS: search control + marker layer + hover tooltip ──────────────────────
search_js <- r"(
function(el, x, data) {
  var map    = this;
  var sites  = data.sites;
  var colors = data.colors;
  var codes  = data.allSpecies;
  var sourceColors = data.sourceColors;
  var sourceList   = data.allSources;
  var mode   = 'species';        // 'species' | 'project'
  var ZOOM_TO_RESULTS = true;   // set false to keep the current view on search

  var home = L.latLngBounds(
    [data.home.lat1, data.home.lng1],
    [data.home.lat2, data.home.lng2]
  );

  var layer = L.layerGroup().addTo(map);

  if (!document.getElementById('pie-marker-style')) {
    var _st = document.createElement('style');
    _st.id = 'pie-marker-style';
    _st.textContent = '.pie-marker{background:transparent;border:none;cursor:pointer;}' +
                      '.pie-marker svg{display:block;}';
    document.head.appendChild(_st);
  }

  // ---- reliable hover tooltip for pie slices (native <title> is flaky) ------
  if (!window._pieTip) {
    var _tip = document.createElement('div');
    _tip.style.cssText = 'position:fixed;z-index:10000;pointer-events:none;' +
      'background:rgba(0,0,0,.82);color:#fff;padding:3px 7px;border-radius:4px;' +
      'font:12px sans-serif;white-space:nowrap;display:none;';
    document.body.appendChild(_tip);
    window._pieTip = _tip;
  }
  var pieTip = window._pieTip;

  if (!el._pieTipBound) {
    el._pieTipBound = true;
    el.addEventListener('mouseover', function(e) {
      var t = e.target;
      var lab = (t && t.getAttribute) ? t.getAttribute('data-label') : null;
      if (lab) {
        pieTip.textContent = lab;
        pieTip.style.left = (e.clientX + 12) + 'px';
        pieTip.style.top  = (e.clientY + 12) + 'px';
        pieTip.style.display = 'block';
      } else {
        pieTip.style.display = 'none';
      }
    });
    el.addEventListener('mousemove', function(e) {
      if (pieTip.style.display === 'block') {
        pieTip.style.left = (e.clientX + 12) + 'px';
        pieTip.style.top  = (e.clientY + 12) + 'px';
      }
    });
  }

  function esc(s) {
    return String(s).replace(/&/g, '&amp;').replace(/</g, '&lt;').replace(/>/g, '&gt;');
  }
  function colorFor(code) { return colors[code] || '#AAAAAA'; }
  function colorForSource(name) { return sourceColors[name] || '#4E79A7'; }

  // ---- click-to-sort for the species tables in pie-marker popups ------------
  window.sortSpeciesTable = function(uid, col, type) {
    var table = document.getElementById('tbl_' + uid);
    if (!table) return;

    var prevCol = table.getAttribute('data-sortcol');
    var prevDir = table.getAttribute('data-sortdir');
    var dir = (prevCol === String(col))
      ? (prevDir === 'asc' ? 'desc' : 'asc')
      : (type === 'num' ? 'desc' : 'asc');   // numbers high->low, text a->z
    table.setAttribute('data-sortcol', col);
    table.setAttribute('data-sortdir', dir);

    var rows = Array.prototype.slice.call(table.getElementsByTagName('tr'));
    var body = rows.filter(function(r) { return !r.getElementsByTagName('th').length; });
    if (!body.length) return;

    body.sort(function(a, b) {
      var av = a.getElementsByTagName('td')[col];
      var bv = b.getElementsByTagName('td')[col];
      av = av ? (av.textContent || av.innerText || '') : '';
      bv = bv ? (bv.textContent || bv.innerText || '') : '';
      var cmp;
      if (type === 'num') {
        cmp = parseFloat(av.replace('%','')) - parseFloat(bv.replace('%',''));
      } else {
        av = av.toUpperCase(); bv = bv.toUpperCase();
        cmp = av < bv ? -1 : (av > bv ? 1 : 0);
      }
      return dir === 'asc' ? cmp : -cmp;
    });

    var parent = body[0].parentNode;
    body.forEach(function(r) { parent.appendChild(r); });
  };

  // Same row order as the tabbed popups: 4-letter codes, other text, then
  // codes starting with a digit; alphabetical within each group.
  function tier(code) {
    if (/^[0-9]/.test(code)) return 3;
    return code.length === 4 ? 1 : 2;
  }
  function bySpeciesCode(a, b) {
    var d = tier(a.code) - tier(b.code);
    if (d !== 0) return d;
    var ua = a.code.toUpperCase(), ub = b.code.toUpperCase();
    return ua < ub ? -1 : (ua > ub ? 1 : 0);
  }

  // ---- unfiltered view: pie icons + tabbed popups ---------------------------
  function showAll(refit) {
    layer.clearLayers();
    if (refit && ZOOM_TO_RESULTS) map.fitBounds(home);
    sites.forEach(function(s) {
      L.marker([s.lat, s.lon], {
        icon: L.divIcon({
          html: s.icon,
          className: 'pie-marker',
          iconSize: [70, 70],
          iconAnchor: [35, 35]
        })
      })
      .bindPopup(s.popup, { maxWidth: 320 })
      .addTo(layer);
    });
    setStatus(sites.length + ' sites, all species');
  }

  // ---- filtered view: SPECIES — one circle per site, years-only popup -------
  function showFilteredSpecies(q) {
    layer.clearLayers();
    var pts = [], nSites = 0, grand = 0;

    sites.forEach(function(s) {
      var hits = s.species.filter(function(sp) {
        return sp.code.toUpperCase().indexOf(q) !== -1;   // use === q for exact
      });
      if (!hits.length) return;
      hits.sort(bySpeciesCode);

      nSites++;
      var tot = hits.reduce(function(a, b) { return a + b.total; }, 0);
      grand += tot;
      pts.push([s.lat, s.lon]);

      var html = '<div style="font-family:sans-serif;min-width:190px"><b>' +
                 esc(s.site) + '</b>';
      hits.forEach(function(sp) {
        html += '<hr style="margin:6px 0"/>' +
          '<b>' + esc(sp.code) + ' &mdash; ' + sp.total + ' detections</b>' +
          '<table style="width:100%;margin-top:4px">' +
          '<tr><th align="left">Year</th><th align="right">n</th></tr>';
        sp.years.forEach(function(yr, i) {
          html += '<tr><td>' + esc(yr) + '</td><td align="right">' + sp.counts[i] + '</td></tr>';
        });
        html += '</table>';
      });
      html += '</div>';

      L.circleMarker([s.lat, s.lon], {
        radius: Math.max(6, Math.min(26, 3.5 * Math.sqrt(tot))),
        color: '#ffffff', weight: 1.5,
        fillColor: colorFor(hits[0].code), fillOpacity: 0.85
      })
      .bindPopup(html, { maxWidth: 320 })
      .bindTooltip(s.site + ' \u2014 ' + tot)
      .addTo(layer);
    });

    if (nSites === 0) {
      setStatus('No sites recorded ' + esc(q) + '. Check the code and try again.');
    } else {
      setStatus(nSites + ' of ' + sites.length + ' sites, ' + grand + ' detections');
      if (ZOOM_TO_RESULTS) map.fitBounds(L.latLngBounds(pts), { padding: [40, 40] });
    }
  }

  // ---- filtered view: PROJECT — one circle per site, species popup ----------
  function showFilteredSource(name) {
    layer.clearLayers();
    var pts = [], nSites = 0, grand = 0;

    sites.forEach(function(s) {
      var src = (s.sources || []).filter(function(x) { return x.name === name; })[0];
      if (!src) return;

      nSites++;
      grand += src.total;
      pts.push([s.lat, s.lon]);

      var html = '<div style="font-family:sans-serif;min-width:190px"><b>' +
                 esc(s.site) + '</b>' +
                 '<hr style="margin:6px 0"/>' +
                 '<b>' + esc(name) + ' &mdash; ' + src.total + ' detections</b>' +
                 '<table style="width:100%;margin-top:4px">' +
                 '<tr><th align="left">Species</th><th align="right">n</th></tr>';
      src.codes.forEach(function(c, i) {
        html += '<tr><td>' + esc(c) + '</td><td align="right">' + src.counts[i] + '</td></tr>';
      });
      html += '</table></div>';

      L.circleMarker([s.lat, s.lon], {
        radius: Math.max(6, Math.min(26, 3.5 * Math.sqrt(src.total))),
        color: '#ffffff', weight: 1.5,
        fillColor: colorForSource(name), fillOpacity: 0.85
      })
      .bindPopup(html, { maxWidth: 320 })
      .bindTooltip(s.site + ' \u2014 ' + src.total)
      .addTo(layer);
    });

    if (nSites === 0) {
      setStatus('No sites recorded for ' + esc(name) + '.');
    } else {
      setStatus(nSites + ' of ' + sites.length + ' sites, ' + grand + ' detections');
      if (ZOOM_TO_RESULTS) map.fitBounds(L.latLngBounds(pts), { padding: [40, 40] });
    }
  }

  function setStatus(txt) {
    var d = el.querySelector('#sp-status');
    if (d) d.innerHTML = txt;
  }

  function run() {
    if (mode === 'species') {
      var q = (el.querySelector('#sp-search').value || '').trim().toUpperCase();
      if (q === '') { showAll(true); } else { showFilteredSpecies(q); }
    } else {
      var name = el.querySelector('#src-select').value || '';
      if (name === '') { showAll(true); } else { showFilteredSource(name); }
    }
  }

  function setMode(m) {
    mode = m;
    var bS = el.querySelector('#mode-species'), bP = el.querySelector('#mode-project');
    var pS = el.querySelector('#panel-species'), pP = el.querySelector('#panel-project');
    // reset both inputs so each toggle starts from "show all"
    el.querySelector('#sp-search').value = '';
    el.querySelector('#src-select').value = '';
    if (m === 'species') {
      bS.style.background = '#4E79A7'; bS.style.color = 'white';
      bP.style.background = '#e0e0e0'; bP.style.color = '#333';
      pS.style.display = ''; pP.style.display = 'none';
    } else {
      bP.style.background = '#4E79A7'; bP.style.color = 'white';
      bS.style.background = '#e0e0e0'; bS.style.color = '#333';
      pP.style.display = ''; pS.style.display = 'none';
    }
    run();
  }

  // ---- the control ----------------------------------------------------------
  var SearchCtl = L.Control.extend({
    onAdd: function() {
      var div = L.DomUtil.create('div', '');
      div.innerHTML =
        '<div style="background:white;padding:9px 12px;border-radius:6px;' +
        'box-shadow:0 1px 5px rgba(0,0,0,.3);font-family:sans-serif;font-size:12px;min-width:190px">' +
          '<div style="display:flex;gap:4px;margin-bottom:7px">' +
            '<button id="mode-species" style="flex:1;padding:4px 6px;border:1px solid #ccc;border-radius:4px;cursor:pointer;font-size:11px;background:#4E79A7;color:white">Species</button>' +
            '<button id="mode-project" style="flex:1;padding:4px 6px;border:1px solid #ccc;border-radius:4px;cursor:pointer;font-size:11px;background:#e0e0e0;color:#333">Project</button>' +
          '</div>' +
          '<div id="panel-species">' +
            '<b>Filter by species</b><hr style="margin:4px 0"/>' +
            '<input id="sp-search" list="sp-codes" autocomplete="off" placeholder="e.g. MYLU" ' +
            'style="width:120px;padding:3px 6px;border:1px solid #ccc;border-radius:4px;font-size:12px"/>' +
            '<button id="sp-clear" style="margin-left:4px;padding:3px 8px;border:1px solid #ccc;' +
            'border-radius:4px;background:#e0e0e0;color:#333;cursor:pointer;font-size:11px">Show all</button>' +
            '<datalist id="sp-codes">' +
            codes.map(function(c) { return '<option value="' + c + '">'; }).join('') +
            '</datalist>' +
          '</div>' +
          '<div id="panel-project" style="display:none">' +
            '<b>Filter by project</b><hr style="margin:4px 0"/>' +
            '<select id="src-select" style="width:130px;padding:3px 6px;border:1px solid #ccc;border-radius:4px;font-size:12px">' +
            '<option value="">Choose\u2026</option>' +
            sourceList.map(function(s) { return '<option value="' + esc(s) + '">' + esc(s) + '</option>'; }).join('') +
            '</select>' +
            '<button id="src-clear" style="margin-left:4px;padding:3px 8px;border:1px solid #ccc;' +
            'border-radius:4px;background:#e0e0e0;color:#333;cursor:pointer;font-size:11px">Show all</button>' +
          '</div>' +
          '<div id="sp-status" style="margin-top:6px;color:#555"></div>' +
        '</div>';
      L.DomEvent.disableClickPropagation(div);
      L.DomEvent.disableScrollPropagation(div);
      return div;
    }
  });
  map.addControl(new SearchCtl({ position: 'topright' }));

  el.querySelector('#mode-species').addEventListener('click', function() { setMode('species'); });
  el.querySelector('#mode-project').addEventListener('click', function() { setMode('project'); });
  el.querySelector('#sp-search').addEventListener('input', run);
  el.querySelector('#sp-clear').addEventListener('click', function() {
    el.querySelector('#sp-search').value = '';
    run();
  });
  el.querySelector('#src-select').addEventListener('change', run);
  el.querySelector('#src-clear').addEventListener('click', function() {
    el.querySelector('#src-select').value = '';
    run();
  });

  showAll(false);   // first draw keeps the extent already set in R
}
)"
  
  # -- Home extent: bounding box of the sites, with a small margin ---------------
  lat_rng <- range(vapply(sites_payload, function(s) s$lat, numeric(1)))
  lng_rng <- range(vapply(sites_payload, function(s) s$lon, numeric(1)))
  pad     <- max(diff(lat_rng), diff(lng_rng)) * 0.05
  pad     <- max(pad, 0.02)   # guard against a single site / zero-width box
  
  home_bounds <- list(
    lat1 = lat_rng[1] - pad, lng1 = lng_rng[1] - pad,
    lat2 = lat_rng[2] + pad, lng2 = lng_rng[2] + pad
  )
  
  m_ok <- leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    fitBounds(home_bounds$lng1, home_bounds$lat1,
              home_bounds$lng2, home_bounds$lat2) %>%
    htmlwidgets::onRender(
      search_js,
      data = list(
        sites        = sites_payload,
        colors       = as.list(species_colors_all),
        allSpecies   = I(all_species_codes),
        sourceColors = as.list(source_colors),
        allSources   = I(all_sources),
        home         = home_bounds
      )
    )
# ── 9. Legend: single-species codes + Myotis & 40KMyo, two columns ────────────
  col_lookup <- if (exists("species_colors_all")) species_colors_all else sp_colors
  
  # Single-species palette codes, split by echolocation frequency
  palette_codes <- setdiff(names(sp_colors), "Other")
  high_codes    <- palette_codes[ grepl("^MY", palette_codes)]   # Myotis  = high freq
  low_codes     <- palette_codes[!grepl("^MY", palette_codes)]   # the rest = low freq
  low_codes     <- palette_codes[!grepl("^MY", palette_codes)]   # the rest = low freq
  
  # Order the low-frequency column blues-first, then greens (by hue)
  low_hsv   <- grDevices::rgb2hsv(grDevices::col2rgb(col_lookup[low_codes]))
  low_codes <- low_codes[order(low_hsv["h", ], decreasing = TRUE)]
  
  # Extra group tags to include, appended to the Myotis (high-freq) column.
  # Names must match the data EXACTLY: "Myotis" (not "MYotis") and "40KMyo".
  extra_tags <- c("Myotis", "40KMyo")
  high_codes <- c(high_codes, extra_tags[extra_tags %in% names(col_lookup)])
  
  legend_row <- function(code) sprintf(
    '<div style="display:flex;align-items:center;margin-bottom:3px"><span style="display:inline-block;width:12px;height:12px;border-radius:50%%;background:%s;margin-right:6px"></span>%s</div>',
    col_lookup[[code]], code)
  
  legend_col <- function(codes) paste(vapply(codes, legend_row, character(1)), collapse = "")
  
  legend_html <- htmltools::HTML(sprintf(
    '<div style="background:white;padding:10px 14px;border-radius:6px;box-shadow:0 1px 5px rgba(0,0,0,.3);font-family:sans-serif;font-size:12px">
     <b>Species</b><hr style="margin:4px 0"/>
     <div style="display:flex;gap:18px">
       <div>%s</div><div>%s</div>
     </div>
   </div>',
   legend_col(high_codes), legend_col(low_codes)
  ))
  
  m_ok <- m_ok %>% addControl(legend_html, position = "bottomright")
```

```{r}
#| label: Set up sites table
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

tbl <- sites |>
  mutate(
    park = factor(park, levels = c(
      "Tplaqin / Enderby Cliffs Park",
      "Kalamalka Lake park",
      "Okanagan Mountain Park",
      "Haynes Lease Ecological Reserve",
      "Skaha Bluffs Park"
    ))
  ) |>
  arrange(park, match(source, c(
    "Sarrell 2012",
    "Paterson 2024",
    "Skurikhina and Kellner 2025"
  ))) |>
  gt(groupname_col = "park") |>
  cols_label(
    site      = "Site",
    lat       = "Latitude",
    lon       = "Longitude",
    y1112     = "2011-12",
    y1819     = "2018-19",
    y1920     = "2019-20",
    y2021     = "2020-21",
    y2122     = "2021-22",
    y2223     = "2022-23",
    y2425     = "2024-25",
    data_type = "Data type",
    source    = html("Project report/data source<br>(BC WSI project ID 6430)")
  ) |>
  tab_spanner(label = "Years monitored",
              columns = c(y1112,y1819, y1920, y2021, y2122, y2223, y2425)) |>
  cols_align(align = "center",
             columns = c(y1112,y1819, y1920, y2021, y2122, y2223, y2425, data_type)) |>
  cols_align(align = "left",  columns = c(site, source)) |>
  cols_align(align = "right", columns = c(lat, lon)) |>
  tab_style(
    style = list(cell_fill(color = "#D5E8F0"), cell_text(weight = "bold")),
    locations = cells_row_groups()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_column_spanners(), cells_column_labels())
  ) |>
  tab_options(
    table.font.size               = px(10),      # smaller for PDF
    data_row.padding              = px(2),
    row_group.padding             = px(3),
    table.border.top.style        = "solid",
    table.border.bottom.style     = "solid",
    column_labels.border.bottom.style = "solid",
    table.width                   = pct(100)     # stretch to full text width
  ) |>
  cols_width(
    site      ~ px(130),
    lat       ~ px(67),
    lon       ~ px(67),
    y1112     ~ px(36),
    y1819     ~ px(36),
    y1920     ~ px(36),
    y2021     ~ px(36),
    y2122     ~ px(36),
    y2223     ~ px(36),
    y2425     ~ px(36),
    data_type ~ px(32),
    source    ~ px(90)
  ) |>
  tab_footnote(
    footnote = "Paterson. 2024. Okanagan winter bat monitoring: 5 year summary 2018-2023. Prepared for Kirk Safford, BC Parks.",
    locations = cells_body(columns = source, rows = source == "Paterson 2024")
  ) |>
  tab_footnote(
    footnote = "Skurikhina and Kellner. 2025. South Okanagan Bat Hibernacula Microsite Pilot Project. 2024-25 field update.",
    locations = cells_body(columns = source, rows = source == "Skurikhina and Kellner 2025")
  )|>
  tab_footnote(
    footnote = "Sarell. 2012. Late Winter Activity of Bats At Skaha Bluffs. Prepared for: BC Parks Penticton, BC",
    locations = cells_body(columns = source, rows = source == "Sarrell 2012")
  )
```

{{< pagebreak >}}

# Land Acknowledgement

Biodiversity Pathways respectfully acknowledges that our work takes place on Treaty 8 and Douglas Treaties Territories as well as the traditional and unceded territories of First Nations and Métis Peoples across all regions of British Columbia, whose histories, languages, and cultures are deeply connected to the biodiversity we monitor. We acknowledge the traditional teachings of the lands that we work on, and that reciprocal, meaningful, and respectful relationships with Indigenous peoples make our work possible. We are deeply grateful for their stewardship of these lands, and we are committed to supporting Indigenous-led monitoring programs, while learning Indigenous ways of knowing, being, and doing.

# Background

Growing evidence indicates that most BC bat species overwinter within the province, with several remaining occasionally active through the winter. This activity has been documented acoustically in both cave and non-cave habitats [@lausen2018], at cliff roosts suspected to be hibernacula in the Peace region (WLRS unpublished data), in coastal forests (WLRS unpublished data), and at five parks in southern BC (BC Parks unpublished data). Further year-round monitoring efforts likely exist beyond those noted here.

Although this activity spans diverse habitats (cliffs, old forests, and waterbodies) little is known about the specific non-cave microsites western bats use in winter [@nagorsen1994; @weller2018]. Documented non-cave roosts include talus and tree roosts in southeast Alaska [@blejwas2021], tree cavities in southeastern BC (deFreitas 2023), a possible talus slope in the same region (Lausen and Hill 2021), and anthropogenic sites such as the siding and trim of houses on the coast (BC Community Bat Program unpublished data). Apart from several large hibernacula in southern BC mines, few hibernacula have been identified, suggesting that most of the region's bats hibernate singly or in small, scattered groups. This mirrors findings from the western US, where wintering Myotis were found singly in 45% of records or in small groups with a median of two, and Townsend's big-eared bats in slightly larger groups with a median of three [@weller2018] .

Understanding how bats use winter habitat has practical value. It can inform the management of recreational and industrial activities, and clarifying how bats use non-cave features such as rock crevices, talus, and trees may improve understanding of how *Pseudogymnoascus destructans* (the fungus that causes white-nose syndrome) spreads in BC, as well as potential treatment options.

Three studies in the Okanagan region, spanning the winter of 2011/12 to the present, form the basis of this report. The first, at Skaha Bluffs in 2011/12, revealed extensive winter activity, recording 1,551 bat passes over 105 sample nights [@sarell2012] . At least eleven species were detected between 7 December 2011 and 31 March 2012 and are therefore thought to hibernate in the area.

BC Parks followed with a multi-year acoustic monitoring study across the Okanagan, running from 2019 through the winter of 2023 to locate potential hibernacula. It recorded ten species in late fall and early winter, and intermittently through the winter, all in rocky outcrops and canyons; this activity was initially interpreted as evidence of hibernacula near the detectors [@paterson2024] . Field surveys in 2024/25, however, pointed to an alternative: bats may have been moving rapidly through these sites, using them as flight corridors rather than roosting nearby. Long-term detector data may therefore be a weaker indicator of local hibernation than first believed, and confirming whether bats actually roost in these features, and pinpointing the microsites they use, would substantially advance understanding of their winter ecology.

This motivated a shift in approach. In 2024/25, @skurikhina2025 established a capture site at Doctor's Wall to locate hibernacula microhabitats directly. The pilot assessed the feasibility of fall and winter capture and tracking in South Okanagan parks and protected areas, and used radiotelemetry to identify the microsites tagged bats used. Specifically, it aimed to:

- assess whether bats could be captured in late fall and early winter in the South Okanagan, particularly Skaha Park and Haynes Lease Ecological Reserve;

- begin identifying the microsites bats use in late fall and winter; and

- test the assumption underlying acoustic monitoring, that late fall and early winter detections indicate bats hibernating nearby, since relocating tagged bats by telemetry would confirm use of the immediate area.

In 2025/26 we also invested in data capture and sharing. All acoustic data from long-term monitoring sites [@paterson2024; @skurikhina2025] were submitted to the North American Bat Monitoring Program database and to WildTrax, supporting larger-scale analyses and allowing the files to be stored and re-analysed in future. Data from @sarell2012 were not added to these databases, but both of the other projects datasets were submitted to the BC Wildlife Species Inventory database.

Finally, we developed a data visualization tool that future results can be added to. Though simple at present, it is designed to be readily expanded into an interactive resource for examining patterns in winter bat detections across Okanagan parks and protected areas.

# Methodology

## Acoustic Monitoring

Passive acoustic monitoring was conducted using Wildlife Acoustic Song Meter MiniBat units, placed out for one to multiple nights, to confirm bat presence and aid in selecting suitable netting microsites. Occasional active monitoring was done during mist-netting, using a handheld Wildlife Acoustic Echometer Touch Pro to monitor bat activity at each capture site. Monitoring effort was not recorded for active monitoring (@tbl-sites).  For the @sarell2012 project acoustic monitoring was done with SM2 Bat and SM2Bat+ units from wildlife acoustics. Most sites were sampled as full-spectrum wave files but the Fortress, Wave, No Climbs and Great White site sin 2012 were sampled only with zero crossing.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: true
#| tbl-cap: "Parks/ Protected Areas, winter acoustic monitoring sites, and years monitored. LT = Long-term monitoring (duration of months); ST = short-term monitoring (duration of days to weeks)." 
#| label: tbl-sites
#| tbl-pos: "H"

tbl
```

Acoustic results for the sites sampled in 2012 were analyzed using Sonobat (version 3.05 Great Basin) if they were full spectrum wave files, and if they were zero-cross files thery were manually reviewd using Analook software. For all other projects data were analyzed using Kaleidoscope Pro (Wildlife Acoustics, version 5.6.8) with bat auto ID Bats of North America 5.8.1 modified for the region. The species list included Pallid Bats (*Antrozous pallidus*, ANPA), Townsend’s Big-eared Bat (*Corynorhinus townsendii*, COTO), Big Brown Bat (*Eptesicus fuscus*, EPFU), Hoary Bat (*Lasiurus cinereus*, LACI), Silver-haired Bat (*Lasionycteris noctivagans*, LANO), Californian Myotis (*Myotis californicus*, MYCA), Western small-footed Myotis (*M. ciliolabrum*, MYCI), Long-eared Myotis (*M. evotis*, MYEV), Little Brown Myotis (*M. lucifugus*, MYLU), Fringed Myotis (*M. thysanodes*, MYTH), Long-legged Myotis (*M. volans*, MYVO), and Yuma Myotis (*M. yumanensis*, MYYU). All files were manually reviewed by Brian Paterson and Mandy Kellner.

## Summary of Results

We summarized results differently depending on the data available for each study. For @skurikhina2025 and @paterson2024 , we worked from the raw detection data; for @sarell2012 , we used the summary data reported in their own study. We then compiled all three datasets into a single dataframe recording the number of species identified per night at each site, along with site location and deployment and retrieval dates. Locations were kept separate even when they shared the same or similar names, so that distinct sites were not conflated in the summaries.

### Data Submission

The long term acoustic monitoring data set for the @paterson2024 and @skurikhina2025 studies has been formatted and submitted to BC Wildlife Species Inventory under project 6430 - South Okanagan Bat Hibernacula Project. Data has been also been submitted to NABat and can be viewed in the [Partner Portal](https://sciencebase.usgs.gov/nabat/#/explore)) under project: South Okanagan Bat Winter Bat Project (Project ID:16855). Analyzed acoustic files have been submitted to [WildTrax](https://portal.wildtrax.ca/discover) and are available can be found under the project: South Okanagan Bat Winter Bat. Access to data sources will require permission from Mandy Kellner, and can be requested directly through the portals.

# Results

Acoustic sampling happened in a total of 7 years, sampling a total of `r length(unique(OK$Sample.Station.Label))` unique sites @fig-monitoring-locations . Across all survey years, `r top_site` had the highest number of recordings in a single year (`r top_site_yr`), with a total of `r formatC(top_site_n, format = "d", big.mark = ",")` detections. The greatest sampling effort in terms of site coverage occurred in `r top_sampled_year`, during which `r top_sampled_nsites` unique sites were monitored.

In total there have been `r nrow(OK)` bat recordings recorded over the length of the study. A total of `r n_confirmed` bat species were confirmed across all sites and years. The most frequently recorded species was `r top_spp_code`, with `r formatC(top_spp_n, format = "d", big.mark = ",")` detections.

::: {.content-visible when-format="pdf"}
To view summary species counts for each site refer to [Appendix A](#appendix-a)
:::

::: {.content-visible when-format="html"}
## Interactive Map
:::

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: true
#| out-width: "100%"
#| fig-align: "center"
#| fig-cap: "Proportion of species bat passes at each site. Only Species or species couplets that identified at least 5% of the total tags appear on map"
#| label: fig-monitoring-locations
if (knitr::pandoc_to("html")) {
  m_ok
} else {
  dir.create(file.path(getwd(), "Figures"), showWarnings = FALSE)
  map_file <- file.path(getwd(), "Figures", "survey_map.png")
  
  # Static legend overlay: addControl legends don't survive the mapshot2 capture,
  # so we inject the same legend_html as a positioned div that IS captured.
  # Sits just above the tile attribution in the bottom-right corner.
  legend_overlay <- htmltools::tags$div(
    style = "position:absolute; bottom:26px; right:8px; z-index:1000;",
    legend_html
  )
  
  m_pdf <- m_ok %>%
    htmlwidgets::prependContent(
      htmltools::tags$style(htmltools::HTML(
        ".leaflet-top.leaflet-right { display:none !important; }"
      )),
      legend_overlay
    )
  
  tryCatch({
    mapview::mapshot2(
      m_pdf,
      file    = map_file,
      delay   = 5,
      vwidth  = 1000,
      vheight = 1350
    )
  }, error = function(e) {
    message("mapshot2 failed: ", e$message)
  })
  
  knitr::include_graphics(map_file)
}
```

# Discussion

There are several underground cave/talus-cave systems known to BC Parks. Assessment of bat use, species presence, and timing of use could be accomplished through acoustic monitoring and/or guano collection. One such area is at Doctor’s wall; during this survey, nets were set up near the entrances of identified caves, but no bats were detected during emergence or caught in the nets. One of the project partners set up an experimental acoustic detector inside the larger cave on November 12, 2024. As of March 2025, this detector had not been retrieved. Options for future work include additional placement of bat detectors at openings or inside caves, single-species DNA analysis through collection of single pellets of guano in paper envelopes (Wildlife Genetics International), and/or multi-species DNA analysis through collection of multiple guano samples (Northern Arizona University Species from Feces lab).

## Recommendations

1.  Explore use of detector array to increase understanding of bat winter movements and potential travel corridors as well as inform capture efforts.
2.  Target Gilles Creek and the Spring for additional monitoring and capture work as the area currently believed to offer the best chance of successful captures.
3.  Increase understanding of bat use of cave/talus features that are known or suspected to be used by bats, through passive acoustic monitoring and guano collection to determine timing of use and species.
4.  Expand collection of existing data. This project aims to identify and include additional acoustic datasets. Possible sources including data from 2025/26 Skaha Park winter monitoring (Kellner pers comm), and ECCC/CWS year-round monitoring at Vaseaux (T. Luszcz pers comm)

# Acknowledgements

This work was supported by funding from the BC Parks License Plate Program, as well as in-kind staff time, vehicles, and equipment from BC Parks and WLRS.

Data included in this compilation and summary was from work done for BC Parks by Brian Paterson and BC Water, Land and Resource Stewardship. Reports and summarized data are available on the BC Wildlife Species Inventory website.

{{< pagebreak >}}

```{r}
#| label: AppendixA
#| echo: false
#| warning: false
#| message: false
#| output: asis
# ---- Appendix A: per-site popup content, PDF only ------------------------
# Recreates the pie chart + "Mean detections by date" chart from the map
# popups, screenshots each with webshot2, one site per subsection.
if (!knitr::pandoc_to("html")) {
  
  appx_dir <- "Figures/AppendixA"
  if (!dir.exists(appx_dir)) dir.create(appx_dir, recursive = TRUE)
  
  pie_min_share <- if (exists("pie_min_share")) pie_min_share else 0.02
  
  # --- Source display order. EDIT to match your exact (trimmed) Source strings.
  #     Anything not listed here sorts alphabetically after these.
  source_order <- c("Sarrell 2012", "Paterson 2024", "Skurikhina/Kellner 2025")
  src_priority <- function(s) {
    p <- match(s, source_order)
    ifelse(is.na(p), length(source_order) + 1L, p)
  }
  
  # --- One row per site: coords, its source(s), and an ordering key -----------
  site_meta <- map_data %>%
    dplyr::mutate(src = trimws(Source)) %>%
    dplyr::group_by(site_name) %>%
    dplyr::summarise(
      lat         = dplyr::first(lat),
      lon         = dplyr::first(lon),
      sources_str = { u <- unique(src); paste(u[order(src_priority(u), u)], collapse = ", ") },
      order_key   = min(src_priority(unique(src))),
      .groups     = "drop"
    ) %>%
    dplyr::arrange(order_key, site_name)   # sites grouped/ordered by source
  
  # --- Build one standalone HTML card (pie + table) per site ------------------
  html_paths <- character(nrow(site_meta))
  card_pngs  <- character(nrow(site_meta))
  
  for (k in seq_len(nrow(site_meta))) {
    site <- site_meta$site_name[k]
    uid  <- gsub("[^a-zA-Z0-9]", "_", site)
    sp   <- map_data %>% dplyr::filter(site_name == site)
    
    # -- pie: identical 2%-pooling logic to the map marker --------------------
    sp_pie_all <- sp %>%
      dplyr::group_by(species_pie) %>%
      dplyr::summarise(n = sum(n), .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(n)) %>%
      dplyr::rename(species = species_pie)
    
    site_total_pie <- sum(sp_pie_all$n)
    keep_species <- sp_pie_all %>%
      dplyr::filter(species != "Other", n / site_total_pie >= pie_min_share) %>%
      dplyr::pull(species)
    
    sp_pie <- sp_pie_all %>%
      dplyr::mutate(species = dplyr::if_else(species %in% keep_species, species, "Other")) %>%
      dplyr::group_by(species) %>%
      dplyr::summarise(n = sum(n), .groups = "drop") %>%
      dplyr::arrange(dplyr::desc(n))
    
    pie_colors <- setNames(
      vapply(sp_pie$species, function(s)
        if (s %in% names(species_colors_all)) species_colors_all[[s]] else "#AAAAAA",
        character(1)),
      sp_pie$species)
    
    pie_svg <- make_pie_svg(sp_pie$n, pie_colors[sp_pie$species],
                            labels = sp_pie$species,
                            color_lookup = species_colors_all,
                            uid = uid, size = 200)   # bigger than the 70px marker
    
    # -- all-years species table (same content as the popup's default tab) ----
    sp_all_table <- sp %>%
      dplyr::group_by(species_table) %>%
      dplyr::summarise(n = sum(n), .groups = "drop") %>%
      dplyr::rename(species = species_table)
    
    table_html <- paste0(
      '<table class="sp"><tr><th>Species</th><th>n</th><th>%</th></tr>',
      make_sp_table(sp_all_table),
      '</table>')
    
    page_html <- sprintf(
      '<!DOCTYPE html><html><head><meta charset="utf-8"><style>
body{margin:0;padding:0;background:#fff;font-family:sans-serif;}
#card{display:inline-flex;align-items:flex-start;gap:22px;padding:14px 18px;background:#fff;}
.pie{flex:0 0 auto;}
.tbl{flex:0 0 auto;}
table.sp{border-collapse:collapse;font-size:13px;min-width:240px;}
table.sp th,table.sp td{border-bottom:1px solid #ddd;padding:3px 10px;text-align:right;}
table.sp th:first-child,table.sp td:first-child{text-align:left;}
table.sp th{border-bottom:2px solid #999;}
</style></head><body>
<div id="card"><div class="pie">%s</div><div class="tbl">%s</div></div>
</body></html>',
      pie_svg, table_html)
    
    safe          <- gsub("[^A-Za-z0-9]+", "_", site)
    html_paths[k] <- normalizePath(file.path(appx_dir, paste0(safe, ".html")), mustWork = FALSE)
    card_pngs[k]  <- file.path(appx_dir, paste0(safe, "_card.png"))
    writeLines(page_html, html_paths[k])
  }
  
  # --- One headless-Chrome session for all cards. asis leaks webshot2's
  #     progress lines into the doc, so capture + discard them. -------------
  invisible(utils::capture.output(suppressMessages(
    webshot2::webshot(url = html_paths, file = card_pngs, selector = "#card", zoom = 3)
  )))
  
  # --- Emit the appendix as markdown ------------------------------------------
  cat("# Appendix A {#appendix-a}\n\n")
  cat("Species detection summaries for each monitoring site.\n\n")
  for (k in seq_len(nrow(site_meta))) {
    if (k > 1) cat("\n\\newpage\n\n")
    cat(sprintf("### %s {.unnumbered .unlisted}\n\n", site_meta$site_name[k]))
    cat(sprintf("**Coordinates:** %.5f, %.5f  \n", site_meta$lat[k], site_meta$lon[k]))
    cat(sprintf("**Source:** %s\n\n", site_meta$sources_str[k]))
    cat(sprintf("![](%s){width=70%%}\n\n", card_pngs[k]))
  }
}
```

{{< pagebreak >}}

# Literature Cited
